---
title: "HANES Project"
output:
  html_document:
    df_print: paged
---

#### Analysis of CDC's 2017-2018 National Health and Nutrition Examination Survey (HANES)

```{r "setup", include=FALSE}
# Set location of Rmd as root dir
knitr::opts_knit$set(root.dir = ".") 
```

Read transformed data subset from disk, and view a random sample of its contents. The complete raw data may be accessed [here](https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2017). Key demographic and nutrition metrics were extracted, and only data from participants providing two days' worth of data (some only participated for a single day) were persisted. Each row represents a food item that was consumed by a participant in a given `day`, and participants are identified by the `id` column.
```{r}
hanes <- read.csv("data/hanes-survey_2017-2018.csv", header=TRUE)
hanes[sample(1:dim(hanes)[1], 10),]
```

Perform one-way ANOVAs between each categorical demographic attribute and each continuous nutrition attribute. Save the p-values and a textual record of the comparisons made.  
```{r}
demographic_columns <- 3:8
nutrition_columns <- 9:length(hanes)

anova_pvalues <- vector()
comparison_labels <- vector()

demographic_counter <- 1

for (demographic_col in hanes[,demographic_columns]) {
  
  nutrition_counter <- 1
  
  for (nutrition_col in hanes[,nutrition_columns]) {
    
    anova_pvalues <- c(anova_pvalues, summary(aov(nutrition_col~demographic_col))[[1]][["Pr(>F)"]][1]) 
    
    comparison_labels <- c(comparison_labels, 
                           paste(colnames(hanes)[nutrition_columns[nutrition_counter]], 
                                 " vs", 
                                 colnames(hanes)[demographic_columns[demographic_counter]]))
    
    nutrition_counter <- nutrition_counter + 1
  }
  demographic_counter <- demographic_counter + 1
}
head(anova_pvalues)
head(comparison_labels)
```

Identify the comparisons that suggest categorical differences.
```{r}
comparison_labels[p.adjust(anova_pvalues, "bonferroni") < 0.05]
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
