---
title: "Demographics, Nutrition, and Health in the United States"
output:
  html_document:
    df_print: paged
---

### Introduction
Proper nutrition is a major factor contributing to human health and longevity. While the dietary quality of most Americans is still poor, it has improved over the years, resulting in substantial decreases to the nation's disease burden [[1](https://www.healthaffairs.org/doi/full/10.1377/hlthaff.2015.0640)]. How this improvement expands across the different groups within the population has not been consistent, and this analysis aims to examine the most recent national nutrition data to determine the relationships between nutrition, health, and demographics. The CDC's National Health and Nutrition Examination Survey for the year of 2017-2018 was used for this purpose. 

This analysis established that lower BMI among non-Hispanic Asians may be explained by this group's lower sugar intake, when compared to non-Hispanic Blacks, Mexican Americans, and non-Hispanic Whites and that non-Hispanic Blacks may be more sensitive to the weight-impacting effects of sugar consumption. Furthermore, elevated BMI was shown to be associated with type 2 diabetes and congestive heart failure, which suggests that non-Hispanic Blacks may be at a higher risk for these diseases. Caffeine consumption was found to be elevated among individuals suffering from type 2 diabetes, coronary heart disease, heart attacks, and strokes, however, it is unclear whether the potential negatives of caffeine consumption outweigh its benefits, which are outlined in the discussion. 

```{r "setup", include=FALSE}
# Set location of Rmd as root dir
knitr::opts_knit$set(root.dir = ".")
# Load libraries
library(tidyverse)
library(rstatix)
library(ggpubr)
```

### Methods

#### Data

The data used in this analysis was obtained from the *Demographic*, *Total Nutrient Intake*, *Body Measures*, *Diabetes* (type 2), and *Medical Conditions* datasets of the CDC's 2017-2018 National Health and Nutrition Examination Survey (NHANES). Each of these documents as well as their corresponding documentation, which describes the attribute units and categorical mappings, may be found [here](https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2017). The data files were imported into a MYSQL database, transformed according to the script available [here](https://github.com/UnexceptedSpectic/adv_binf_stats/blob/main/notebooks/project/data/etl.sql). Key demographic, health, and nutrition metrics were extracted and combined into a single table to facilitate statistical analysis. This combined table was exported as a *csv*, contains 15,656 records, and is available [here](https://github.com/UnexceptedSpectic/adv_binf_stats/blob/main/notebooks/project/data/nhanes-survey_2017-2018.csv). Apart from the `id` column, the values in each row represent either demographic/health categorical factors or continuous values representing total daily nutrient intakes (in grams, apart from `total_kcal`, which is in kilocalories) for an individual participating in the survey. The database containing the raw and combined data may be recreated by using [this](https://github.com/UnexceptedSpectic/adv_binf_stats/blob/main/notebooks/project/data/db_dump.sql) SQL dump file. 

```{r include=FALSE}
hanes <- read.csv("data/nhanes-survey_2017-2018.csv", header=TRUE)
hanes[sample(1:dim(hanes)[1], 10),]
```

#### Statistical Models
R was used to perform ANOVAs between each categorical (demographic and health) variable and each continuous (nutrition data) variable to identify variables that are categorically different with respect to their continuous counterparts. Pairwise t-tests were then performed on each of these variables to determine exactly which categories differed between each other in their means. For both tests, Bonferroni correction was used to adjust the resulting p-values. All data and additional protocol details are available [here](https://github.com/UnexceptedSpectic/adv_binf_stats/tree/main/notebooks/project). 

```{r include=FALSE}
# Remove pregnant individuals from consideration, to prevent confounding effects. 1 indicates pregnant

hanes <- hanes[(hanes$pregnant != 1),]
```

```{r include=FALSE}
# Define data columns as those having less than 10 unique values - an arbitrary, but manually verified cutoff value.

category_logical <- lapply(hanes, function(x) length(unique(x)) < 10)
categorical_columns <- (1:length(hanes))[unlist(category_logical)]
continuous_columns <- (1:length(hanes))[!unlist(category_logical)]
```

```{r include=FALSE}
# Perform one-way ANOVAs between each categorical variable and each continuous variable, to determine which comparisons suggest significant categorical differences. Save the p-values and a textual record of the comparisons made.  

anova_pvalues <- vector()
comparison_columns <- list()

comparison_counter <- 1
category_counter <- 1

for (cat_col in hanes[,categorical_columns]) {
  
  continuous_counter <- 1
  
  for (cont_col in hanes[,continuous_columns]) {
    
    # Zeros represent missing values, so they are filtered out from the continuous variables
    anova_pvalues <- c(anova_pvalues, summary(aov(cont_col[cont_col!=0]~cat_col[cont_col!=0]))[[1]][["Pr(>F)"]][1]) 
    
    comparison_columns[[comparison_counter]] <- c(categorical_columns[category_counter], continuous_columns[continuous_counter])
    
    continuous_counter <- continuous_counter + 1
    comparison_counter <- comparison_counter + 1
  }
  category_counter <- category_counter + 1
}
head(anova_pvalues)
head(comparison_columns, 1)
```

```{r include=FALSE}
# Identify and count the comparisons that suggest categorical differences.

significant_comparisons <- p.adjust(anova_pvalues, "bonferroni") < 0.05
sum(significant_comparisons)
```

```{r include=FALSE}
# Define a function for obtaining continuous variable means, grouped by categorical factors. 

getTtestGroupMeans <- function(test, group, cont_col, cat_col) {
  means <- vector()
  groups <- vector()
  if (group == 1) {
    groups <- test$group1
  } else if (group == 2) {
    groups <- test$group2
  } else {
    print("Error: Group must be 1 or 2, as this function takes a t-test result as input")
    return()
  }
  for (i in groups) {
    means <- c(means, round(mean(cont_col[cat_col == i]), 2))
  }
  return(means)
}
```

```{r include=FALSE}
# Define a function for executing pairwise t-tests between each continuous and categorical variable.
# Rows whose continuous variable has a zero value are filtered out.

pw_ttest <- function(comp_col_pair, cat_col, cont_col) {
  data = tibble(cat = factor(cat_col[cont_col!=0]), cont = cont_col[cont_col!=0])
  stat.test <- data %>%
    pairwise_t_test(
      cont ~ cat,
      p.adjust.method="bonferroni"
      )
  return(stat.test %>% 
         add_column(cat_ind=comp_col_pair[1]) %>% 
         add_column(cont_ind=comp_col_pair[2]) %>% 
         add_column(comp=paste(colnames(hanes)[comp_col_pair[1]], "vs", colnames(hanes)[comp_col_pair[2]])) %>%
         add_column(x1=getTtestGroupMeans(test=stat.test, group=1, cat_col=cat_col[cont_col!=0], cont_col=cont_col[cont_col!=0])) %>%
         add_column(x2=getTtestGroupMeans(test=stat.test, group=2, cat_col=cat_col[cont_col!=0], cont_col=cont_col[cont_col!=0]))
  )
}
```

```{r include=FALSE}
# Perform pairwise t-tests between each category for variables found to have significant categorical differences. 

t_tests <- NULL
for (comp_col_pair in comparison_columns[significant_comparisons]) {
  t_test <- pw_ttest(comp_col_pair=comp_col_pair,cat_col=hanes[,comp_col_pair[1]], cont_col=hanes[,comp_col_pair[2]])
  t_tests <- bind_rows(t_tests, t_test)
}
```

```{r include=FALSE}
#Isolate comparisons showing significant differences. 

sig_t_tests <- t_tests[t_tests$p.adj < 0.05,]
sig_t_tests <- sig_t_tests[order(sig_t_tests$p.adj),]
# Groups with 0 values represent missing information - questions unanswered by participants 
sig_t_tests <- sig_t_tests[((sig_t_tests$group1 != 0) & (sig_t_tests$group2 != 0)),]
# Subset and rearrange columns
sig_t_tests <- tibble(
  comparison=sig_t_tests$comp,
  cat_ind=sig_t_tests$cat_ind,
  cont_ind=sig_t_tests$cont_ind,
  group1=sig_t_tests$group1, 
  group2=sig_t_tests$group2, 
  group1.mean=sig_t_tests$x1, 
  group2.mean=sig_t_tests$x2, 
  group1.size=sig_t_tests$n1,
  group2.size=sig_t_tests$n2,
  pval.adj=sig_t_tests$p.adj)
head(sig_t_tests)
dim(sig_t_tests)
```

### Results and Conclusions

```{r include=FALSE}
# Define helper functions for subsetting and annotating t-tests.

hanesColInd <- function(colName) {
  return((1:length(hanes))[colnames(hanes) == colName])
}

getTtest <- function(colname, cat_labels) {
  res <- sig_t_tests[(sig_t_tests$cat_ind==hanesColInd(colname)),]
  res$group1 <- cat_labels[as.numeric(res$group1)]
  res$group2 <- cat_labels[as.numeric(res$group2)]
  res <- cbind(res[,1], res[,(4:length(res))])
  return(res)
}
```

#### Dietary and health trends by race
Dietary and health metric disparities across race were screened for and Table 1 shows a set of significant findings and their associated p-values. We show that non-Hispanic Asians tend to have lower BMI than non-Hispanic Blacks, Mexican Americans, and non-Hispanic Whites. This may be explained by our analogous finding that non-Hispanic Asians tend to consume less sugar than non-Hispanic Blacks, Mexican Americans, and non-Hispanic Whites. 

Furthermore, non-Hispanic Blacks were shown to have higher BMI than non-Hispanic Whites. The only difference in nutritional intake between non-Hispanic Blacks and other racial groups was the finding that non-Hispanic Whites tend to consume more sugar than do non-Hispanic Blacks and Mexican Americans.

**Table 1 | Comparisons of BMI and sugar intake across racial categories.** The categorical and continuous variables used in a particular pairwise t-test are described in the `comparison` column. The specific racial groups compared are detailed in the columns `group1` and `group2`, and each group's mean and sample size is detailed in the `groupX.mean` and `groupX.size` columns. BMI is given in (weight (kg)/height (m)<sup>2</sup>) and sugar intake is given in grams/day. The Bonferroni adjusted p-values associated with each t-test are shown in the the `pval.adj` column. All shown results are significant at p < 0.05.
```{r echo=FALSE}
race_table <- getTtest(colname="race", cat_labels=c("Mexican American", "Other Hispanic", "non-Hispanic White", "non-Hispanic Black", "", "non-Hispanic Asian", "Other"))
race_table <- race_table[((race_table$group2 != "Other") & (race_table$group1 !="Other Hispanic")),]
race_table <- race_table[grep("bmi$|sugar$", race_table$comparison),]
remove_rownames(race_table[order(race_table$comparison),])
```

<br>

```{r include=FALSE}
perc_diff <- function(f, i) {return(round((f-i)/i*100, 0))}
```


#### Diseases Linked to High BMI and Elevated Caffeine Intake
BMI and caffeine intake were compared across disease status and the results, as well as their their associated p-values, are shown in Table 2. We found that individuals suffering from type 2 diabetes tend to have higher BMI than do those that don't have the disease. Likewise, individuals with congestive heart failure tend to have higher BMI than do individuals healthy in this respect. 

We also found that those suffering from type 2 diabetes, coronary heart disease, heart attacks, and strokes all consume higher levels of caffeine than do individuals without these diseases. The amounts by which caffeine consumption in diseased individuals exceeds that of healthy individuals are `r perc_diff(157.11, 126.08)`%, `r perc_diff(195.62, 164.62)`%, `r perc_diff(244.32, 161.81)`%, and `r perc_diff(208.73, 163.97)`%, respectively.

**Table 2 | Comparisons of BMI and caffeine intake across disease status.** The categorical variables representing diseases (type 2 diabetes, coronary heart disease, heart attack, stroke) and continuous variables (BMI and caffeine) used in a particular pairwise t-test are described in the `comparison` column. The disease status of the groups compared is detailed in the columns `group1` and `group2`, and each group's mean and sample size is detailed in the `groupX.mean` and `groupX.size` columns. BMI is given in (weight (kg)/height (m)<sup>2</sup>) and caffeine intake is given in grams/day. The Bonferroni adjusted p-value associated with each t-test is shown in the the `pval.adj` column. All shown results are significant at p < 0.05.
```{r echo=FALSE}
diabetes_table <- getTtest(colname="diabetes", cat_labels=c("diseased", "healthy", "borderline"))
chf_table <- getTtest(colname="cong_heart_failure", cat_labels=c("diseased", "healthy", rep("", 6), "Don't Know"))
chd_table <- getTtest(colname="cor_heart_disease", cat_labels=c("diseased", "healthy", rep("", 6), "Don't Know"))
ha_table <- getTtest(colname="had_heart_attack", cat_labels=c("diseased", "healthy", rep("", 6), "Don't Know"))
stroke_table <- getTtest(colname="had_stroke", cat_labels=c("diseased", "healthy"))

combined_table <- rbind(
  diabetes_table[grep("bmi$|caffeine$", diabetes_table$comparison),],
  chf_table[grep("bmi$|caffeine$", chf_table$comparison),],
  chd_table[grep("bmi$|caffeine$", chd_table$comparison),],
  ha_table[grep("bmi$|caffeine$", ha_table$comparison),],
  stroke_table[grep("bmi$|caffeine$", stroke_table$comparison),]
)

combined_table <- combined_table[((combined_table$group2 != "borderline") & (combined_table$group2 != "Don't Know")),]
remove_rownames(combined_table[order(unlist(lapply(combined_table$comparison, function(x) strsplit(x, 'vs')[[1]][2]))),])
```

### Discussion
While the results showcased above are all significant (p<0.05), it is not necessarily so that one compared variable, for instance a nutrition metric like sugar intake, is directly responsible for a categorical outcome, such as type 2 diabetes. The categorical outcome may have been present prior to the observation of the nutrition value. Therefore, the below discussion is only that of the possible relationships that could exist between the compared variables outlined in the result section. Inferences are supported with existing literature, however, more detailed studies may have to be designed to confirm them. 

#### Dietary and health trends by race

The first interesting insight is that non-Hispanic Asians may have lower BMI than do non-Hispanic Blacks, Mexican Americans, and non-Hispanic Whites due to their lower intake of sugar, all other studied nutrient metrics being equal. One of the best studied risk factors of higher BMI is the consumption of sugar-sweetened beverages [[2](https://academic.oup.com/ajcn/article/84/2/274/4881805?login=true)], and non-Hispanic Asian youths have been shown to consume less sugar-sweetened beverages than do non-Hispanic white, non-Hispanic black, and Hispanic youths [[3](https://stacks.cdc.gov/view/cdc/44039)]. The reason for this observation is not entirely clear and may result from some combination of both genetic predispositions and socioeconomic factors. 

Secondly, non-Hispanic Blacks tend to have higher BMI when compared to non-Hispanic whites despite consuming less sugar, all other studied metrics being equal. This suggests that non-Hispanic Blacks may be more sensitive to the effects of sugar, than are non-Hispanic whites. One study showed that obese African American women are more sensitive to glucose in their insulin response than are obese White American women [[4](https://pubmed.ncbi.nlm.nih.gov/28629856/)]. This, coupled with the facts that obesity is linked to low socioeconomic status [[5](https://www.liebertpub.com/doi/abs/10.1089/chi.2015.0029)] and that racial minorities are more likely to occupy lower socioeconomic strata [[6](https://www.bls.gov/spotlight/2018/race-economics-and-social-status/pdf/race-economics-and-social-status.pdf)], suggest that our observation may be reasonably qualified. 

#### Diseases Linked to High BMI and Elevated Caffeine Intake

Two diseases were found to be associated with higher BMI when compared to respective non-diseased group BMI: type 2 diabetes and congestive heart failure. Given that the majority of individuals diagnosed with type 2 diabetes are overweight or obese [[7](https://pubmed.ncbi.nlm.nih.gov/17889292/)] and that intentional weight loss among individuals with type 2 diabetes reduces mortality [[8](https://care.diabetesjournals.org/content/23/10/1499.short)], it is likely that higher-than-healthy BMI is a risk factor for the development of type 2 diabetes. Similarly, research has shown that obese individuals have nearly twice the risk of developing congestive heart failure as do those with healthy weight [[9](https://www.nejm.org/doi/full/10.1056/NEJMoa020245)] and those suffering from congestive heart failure reap similar benefits by losing weight as do overweight individuals without the disease [[10](https://pubmed.ncbi.nlm.nih.gov/9315579/)]. 

Furthermore, we found that elevated caffeine intake may be associated with type 2 diabetes, coronary heart disease, heart attacks, and strokes. Caffeine consumption has been shown to impair glucose tolerance in men[[11](https://www.sciencedirect.com/science/article/abs/pii/S0140673667903340)] and to decrease insulin sensitivity in young adults [[12](https://care.diabetesjournals.org/content/27/8/2047.short)]. Although the latest research suggests there is no increase in risk of coronary heart disease or stroke with caffeine consumption, conflicting results exist regarding the relationship between caffeine consumption and myocardial infarction [[13](https://www.nejm.org/doi/full/10.1056/NEJM199010113231504)]. While some research suggests that caffeine may be beneficial toward reducing the risk of Alzheimer's disease [[14](https://pubmed.ncbi.nlm.nih.gov/20182037/)] and Parkinson disease [[15](https://pubmed.ncbi.nlm.nih.gov/10819950/)], it may still benefit individuals to consume it in moderation. 
